---
title: "NATSAP- OQ diff by RTCWT"
author: "Sean Raleigh"
date: "August 19, 2015"
output: html_document
---

This document contains the analyis of OQ difference with RTC/WT as a predictor.

## Housekeeping

Load necessary libraries and special functions.

```{r, message = FALSE}
library(rstan)
library(parallel)
library(beepr)
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(stringr)
library(tidyr)
library(lme4)
```

```{r}
source("stan_utilities.R")
```

## Import data

```{r}
natsap_tidy <- read_csv("./tidydata/natsap_tidy.csv")
dose_tidy <- read_csv("./tidydata/dose_tidy.csv")
```


## No pooling

Here we analyze OQ diff by for RTC and WT separately. (Complete pooling here would just be OQdiff, which has already been analyzed.)

Gather data to pass to Stan:

```{r}
# model name is 'OQdiff_RTCWT'

# Grab RTCWT status for each client.
RTCWT <- dose_tidy$RTCWT[natsap_tidy$new_ID]
RTCWT <- ifelse(RTCWT == "RTC", 1, 2)  # 1 = RTC, 2 = WT

data_OQdiff_RTCWT <- list(y = natsap_tidy$OQdiff,
                          n = length(natsap_tidy$OQdiff),
                          RTCWT = RTCWT)
```

Mathematical model:

$$ y_{i} \sim N\left( \mu_{RTCWT[i]}, \sigma_{RTCWT[i]} \right), \quad 1 \leq i \leq n $$

Write Stan model:

```{r}
cat("
    data {
        int<lower = 1> n;                       // clients
        vector<lower = -256, upper = 256>[n] y; // OQ diffs
        int<lower = 1, upper = 2> RTCWT[n];     // RTC/WT
    }
    parameters {
        vector<lower = -256, upper = 256>[2] mu;
        vector<lower = 0>[2] sigma;
    }
    model {
        // Implicit uniform priors on mus and sigmas
        for (i in 1:n) {
            y[i] ~ normal(mu[RTCWT[i]], sigma[RTCWT[i]]);
        }
    }
",
file = "./OQdiff_RTCWT/OQdiff_RTCWT.stan")
```

Fit Stan model:

```{r}
final_model = TRUE

if (file.exists("./OQdiff_RTCWT/OQdiff_RTCWT.rds") & final_model == TRUE) {
    stan_OQdiff_RTCWT <- readRDS("./OQdiff_RTCWT/OQdiff_RTCWT.rds")
} else {
    stan_OQdiff_RTCWT <- stan_model("./OQdiff_RTCWT/OQdiff_RTCWT.stan")
    saveRDS(stan_OQdiff_RTCWT, file = "./OQdiff_RTCWT/OQdiff_RTCWT.rds")
}

fit_OQdiff_RTCWT <- sampling(stan_OQdiff_RTCWT, data = data_OQdiff_RTCWT)
```

Extract and save samples:

```{r}
samples_OQdiff_RTCWT <- as.data.frame(fit_OQdiff_RTCWT)

write_csv(samples_OQdiff_RTCWT, path = "./OQdiff_RTCWT/OQdiff_RTCWT.csv")

params_OQdiff_RTCWT <- c("mu[1]", "mu[2]", "sigma[1]", "sigma[2]")
```

Show chain diagnostics and posterior summaries:

```{r, fig.width = 9}
traceplot(fit_OQdiff_RTCWT, pars = params_OQdiff_RTCWT, inc_warmup = FALSE)

print(fit_OQdiff_RTCWT)

len <- length(params_OQdiff_RTCWT)
sample_plots(samples_OQdiff_RTCWT,
             params = params_OQdiff_RTCWT,
             cred_mass = 0.95,
             layout = matrix(c(1:(len + len %% 2)),
                             nrow = (len + len %% 2)/2,
                             ncol = 2))
```

Clean up workspace:

```{r, results = 'hide'}
cleanup("OQdiff_RTCWT")
rm(final_model, len)
gc()
```


## Partial pooling model

Here we explore what it might look like to use RTCWT as a group-level predictor.

Gather data to pass to Stan:

```{r}
# model name is 'OQdiff_RTCWT_hier'

data_OQdiff_RTCWT_hier <- list(y = natsap_tidy$OQdiff,
                               n = length(natsap_tidy$OQdiff),
                               RTCWT = ifelse(dose_tidy$RTCWT == "RTC", 0, 1),
                                    # 0 = RTC, 1 = WT
                               J = NROW(dose_tidy),
                               prog = natsap_tidy$new_ID)
```

Mathematical model:

$$ \alpha_{j} \sim N\left(\gamma_{\alpha} + 
    \gamma_{WT} RTCWT_{j}, \sigma_{\alpha} \right), \quad 1 \leq j \leq J $$
$$ y_{i} \sim N\left( \alpha_{j[i]}, \sigma_{y} \right), \quad 1 \leq i \leq n $$

Write Stan model:

```{r}
cat("
    data {
        int<lower = 1> n;                       // clients
        vector<lower = -256, upper = 256>[n] y; // OQ diffs
        int<lower = 1> J;                       // number of programs
        int prog[n];                            // programs
        vector<lower = 0, upper = 1>[J] RTCWT;  // RTC/WT
        
    }
    parameters {
        real<lower = -256, upper = 256> gamma_a; // program-level mean
        real<lower = -256, upper = 256> WT;      // coefficient of RTC/WT        
        real<lower = 0> sigma_a;                 // program-level sd
        vector<lower = -256, upper = 256>[J] a;  // program intercepts
        real<lower = 0> sigma_y;                 // data-level sd
    }
    model {
        // Implicit uniform priors on mus and sigmas
        a ~ normal(gamma_a + WT * RTCWT, sigma_a);
        for (i in 1:n) {
            y[i] ~ normal(a[prog[i]], sigma_y);
        }
    }
",
file = "./OQdiff_RTCWT/OQdiff_RTCWT_hier.stan")
```

Fit Stan model:

```{r}
final_model = TRUE

if (file.exists("./OQdiff_RTCWT/OQdiff_RTCWT_hier.rds") & final_model == TRUE) {
    stan_OQdiff_RTCWT_hier <- readRDS("./OQdiff_RTCWT/OQdiff_RTCWT_hier.rds")
} else {
    stan_OQdiff_RTCWT_hier <- stan_model("./OQdiff_RTCWT/OQdiff_RTCWT_hier.stan")
    saveRDS(stan_OQdiff_RTCWT_hier, file = "./OQdiff_RTCWT/OQdiff_RTCWT_hier.rds")
}

fit_OQdiff_RTCWT_hier <- sampling(stan_OQdiff_RTCWT_hier, data = data_OQdiff_RTCWT_hier)
```

Extract and save samples:

```{r}
samples_OQdiff_RTCWT_hier <- as.data.frame(fit_OQdiff_RTCWT_hier)

write_csv(samples_OQdiff_RTCWT_hier, path = "./OQdiff_RTCWT/OQdiff_RTCWT_hier.csv")

# Get 4 random programs
set.seed(1234)
some_programs <- sort(sample(1:data_OQdiff_RTCWT_hier$J, 4))

# Display sample sizes for selected programs.
cat("Sample sizes for programs", some_programs, ": ",
    dose_tidy$n[dose_tidy$new_ID %in% some_programs])

# Gather up parameter names
alphas <- some_programs %>%
    as.character() %>%
    sapply(function(x) {paste("a[", x, "]", sep = "")}) %>%
    unname()

params_OQdiff_RTCWT_hier <- c("gamma_a", "sigma_a", "WT", "sigma_y", alphas)
```

Show chain diagnostics and posterior summaries:

```{r, fig.width = 9, fig.height = 10}
traceplot(fit_OQdiff_RTCWT_hier, pars = params_OQdiff_RTCWT_hier, inc_warmup = FALSE)

print(fit_OQdiff_RTCWT_hier)

len <- length(params_OQdiff_RTCWT_hier)
sample_plots(samples_OQdiff_RTCWT_hier,
             params = params_OQdiff_RTCWT_hier,
             cred_mass = 0.95,
             layout = matrix(c(1:(len + len %% 2)),
                             nrow = (len + len %% 2)/2,
                             ncol = 2, byrow = TRUE))
```

Clean up workspace:

```{r, results = 'hide'}
cleanup("OQdiff_RTCWT_hier")
rm(final_model, len, some_programs, alphas)
gc()
```

```{r}
beep()
```
