---
title: "NATSAP- OQ diff by RTCWT"
author: "Sean Raleigh"
date: "July 12, 2015"
output: html_document
---

This document contains the analyis of OQ difference with RTC/WT as a predictor.

## Housekeeping

Load necessary libraries and special functions.

```{r, message = FALSE}
library(rstan)
library(parallel)
library(beepr)
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(stringr)
library(tidyr)
library(lme4)
```

```{r}
source("stan_utilities.R")
```

## Import data

```{r}
natsap_tidy <- read_csv("./tidydata/natsap_tidy_1.csv")
dose_tidy <- read.csv("./tidydata/dose_tidy_1.csv")
```


## No pooling

Here we analyze OQ diff by for RTC and WT separately. (Complete pooling here would just be OQdiff, which has already been analyzed.)

Gather data to pass to Stan:

```{r}
# model name is 'OQdiff_RTCWT'

# Grab RTCWT status for each client.
RTCWT <- dose_tidy$RTCWT[natsap_tidy$new_ID]
RTCWT <- ifelse(RTCWT == "RTC", 1, 2)  # 1 = RTC, 2 = WT

data_OQdiff_RTCWT <- list(y = natsap_tidy$OQdiff,
                          n = length(natsap_tidy$OQdiff),
                          RTCWT = RTCWT)
```

Write Stan model:

```{r}
cat("
    data {
        int<lower = 1> n;                       // clients
        vector<lower = -256, upper = 256>[n] y; // OQ diffs
        int<lower = 1, upper = 2> RTCWT[n];     // RTC/WT
    }
    parameters {
        vector<lower = -256, upper = 256>[2] mu;
        vector<lower = 0>[2] sigma;
    }
    model {
        // Implicit uniform priors on mus and sigmas
        for (i in 1:n) {
            y[i] ~ normal(mu[RTCWT[i]], sigma[RTCWT[i]]);
        }
    }
",
file = "./OQdiff_RTCWT/OQdiff_RTCWT.stan")
```

Fit Stan model:

```{r}
final_model = FALSE

if (file.exists("./OQdiff_RTCWT/OQdiff_RTCWT.rds") & final_model == TRUE) {
    stan_OQdiff_RTCWT <- readRDS("./OQdiff_RTCWT/OQdiff_RTCWT.rds")
} else {
    stan_OQdiff_RTCWT <- stan_model("./OQdiff_RTCWT/OQdiff_RTCWT.stan")
    saveRDS(stan_OQdiff_RTCWT, file = "./OQdiff_RTCWT/OQdiff_RTCWT.rds")
}

fit_OQdiff_RTCWT <- sampling(stan_OQdiff_RTCWT, data = data_OQdiff_RTCWT)
```

Extract and save samples:

```{r}
samples_OQdiff_RTCWT <- as.data.frame(fit_OQdiff_RTCWT)

write_csv(samples_OQdiff_RTCWT, path = "./OQdiff_RTCWT/OQdiff_RTCWT.csv")

params_OQdiff_RTCWT <- c("mu[1]", "mu[2]", "sigma[1]", "sigma[2]")
```

Show chain diagnostics and posterior summaries:

```{r, fig.width = 9}
traceplot(fit_OQdiff_RTCWT, pars = params_OQdiff_RTCWT, inc_warmup = FALSE)

print(fit_OQdiff_RTCWT)

len <- length(params_OQdiff_RTCWT)
sample_plots(samples_OQdiff_RTCWT,
             params = params_OQdiff_RTCWT,
             cred_mass = 0.95,
             layout = matrix(c(1:(len + len %% 2)),
                             nrow = (len + len %% 2)/2,
                             ncol = 2))
```

Clean up workspace:

```{r, results = 'hide'}
cleanup("OQdiff_RTCWT")
rm(final_model, len)
gc()
```

```{r}
beep()
```
