---
title: "NATSAP- OQ diff by sex"
author: "Sean Raleigh"
date: "July 11, 2015"
output: html_document
---

This document contains the analyis of OQ difference with sex as a predictor.

## Housekeeping

Load necessary libraries and special functions.

```{r, message = FALSE}
library(rstan)
library(parallel)
library(beepr)
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(stringr)
library(tidyr)
library(lme4)
```

```{r}
source("stan_utilities.R")
```

## Import data

```{r}
natsap_tidy <- read_csv("./tidydata/natsap_tidy_1.csv")
dose_tidy <- read.csv("./tidydata/dose_tidy_1.csv")
```


## Complete pooling model

Here we analyze OQdiff by sex (averaged over all programs).

Gather data to pass to Stan:

```{r}
# model name is 'OQdiff_sex'

data_OQdiff_sex <- list(y = natsap_tidy$OQdiff,
                        n = length(natsap_tidy$OQdiff),
                        sex = ifelse(natsap_tidy$sex == "MALE", 0, 1))
```

Write Stan model:

```{r}
cat("
    data {
        int<lower = 1> n;                       // clients
        vector<lower = -256, upper = 256>[n] y; // OQ diffs
        vector<lower = 0, upper = 1>[n] sex;    // sex
    }
    parameters {
        real intercept;                               // intercept
        real female;                             // coefficient of sex
        real<lower = 0> sigma;
    }
    model {
        // Implicit uniform priors on intercept, female, and sigma
        y ~ normal(intercept + female * sex, sigma);
    }
",
file = "./OQdiff_sex/OQdiff_sex.stan")
```

Fit Stan model:

```{r}
final_model = TRUE

if (file.exists("./OQdiff_sex/OQdiff_sex.rds") & final_model == TRUE) {
    stan_OQdiff_sex <- readRDS("./OQdiff_sex/OQdiff_sex.rds")
} else {
    stan_OQdiff_sex <- stan_model("./OQdiff_sex/OQdiff_sex.stan")
    saveRDS(stan_OQdiff_sex, file = "./OQdiff_sex/OQdiff_sex.rds")
}

fit_OQdiff_sex <- sampling(stan_OQdiff_sex, data = data_OQdiff_sex)
```

Extract and save samples:

```{r}
samples_OQdiff_sex <- as.data.frame(fit_OQdiff_sex)

write_csv(samples_OQdiff_sex, path = "./OQdiff_sex/OQdiff_sex.csv")

params_OQdiff_sex <- c("intercept", "female", "sigma")
```

Show chain diagnostics and posterior summaries:

```{r, fig.width = 9}
traceplot(fit_OQdiff_sex, pars = params_OQdiff_sex, inc_warmup = FALSE)

print(fit_OQdiff_sex)

len <- length(params_OQdiff_sex)
sample_plots(samples_OQdiff_sex,
             params_OQdiff_sex,
             cred_mass = 0.95,
             layout = matrix(c(1:(len + len %% 2)),
                             nrow = (len + len %% 2)/2,
                             ncol = 2))
```

Clean up workspace:

```{r, results = 'hide'}
cleanup("OQdiff_sex")
rm(final_model, len)
gc()
```

## No pooling model

Now we break it down by program. In other words, the following produces a regression of OQdiff by sex for each program separately.

Gather data to pass to Stan:

```{r}
# model name is 'OQdiff_sex_prog'

data_OQdiff_sex_prog <- list(y = natsap_tidy$OQdiff,
                           n = length(natsap_tidy$OQdiff),
                           sex = ifelse(natsap_tidy$sex == "MALE", 0, 1),
                           J = NROW(dose_tidy),
                           prog = natsap_tidy$new_ID)
```

Write Stan model:

```{r}
cat("
    data {
        int<lower = 1> n;                       // clients
        vector<lower = -256, upper = 256>[n] y; // OQ diffs
        vector<lower = 0, upper = 1>[n] sex;    // sex
        int<lower = 1> J;                       // programs
        int prog[n];                            // program for each client
    }
    parameters {
        real female;                              // coefficient of sex
        vector<lower = -256, upper = 256>[J] mu; // program means for males
        vector<lower = 0>[J] sigma;              // program sds
    }
    model {
        // Implicit uniform priors on female, mus, and sigmas
        for (i in 1:n) {
            y[i] ~ normal(mu[prog[i]] + female * sex[i], sigma[prog[i]]);
        }
    }
",
file = "./OQdiff_sex/OQdiff_sex_prog.stan")
```

Fit Stan model:

```{r}
final_model = TRUE

if (file.exists("./OQdiff_sex/OQdiff_sex_prog.rds") & final_model == TRUE) {
    stan_OQdiff_sex_prog <- readRDS("./OQdiff_sex/OQdiff_sex_prog.rds")
} else {
    stan_OQdiff_sex_prog <- stan_model("./OQdiff_sex/OQdiff_sex_prog.stan")
    saveRDS(stan_OQdiff_sex_prog, file = "./OQdiff_sex/OQdiff_sex_prog.rds")
}

fit_OQdiff_sex_prog <- sampling(stan_OQdiff_sex_prog, data = data_OQdiff_sex_prog)
```

Extract and save samples:

```{r}
samples_OQdiff_sex_prog <- as.data.frame(fit_OQdiff_sex_prog)

write_csv(samples_OQdiff_sex_prog, path = "./OQdiff_sex/OQdiff_sex_prog.csv")

# Get 4 random programs
set.seed(1234)
some_programs <- sort(sample(1:data_OQdiff_sex_prog$J, 4))

# Display sample sizes for selected programs.
cat("Sample sizes for programs", some_programs, ": ",
    dose_tidy[dose_tidy$new_ID %in% some_programs,"n"])

# Gather up parameter names
mus <- some_programs %>%
    as.character() %>%
    sapply(function(x) {paste("mu[", x, "]", sep = "")}) %>%
    unname()
sigmas <- some_programs %>%
    as.character() %>%
    sapply(function(x) {paste("sigma[", x, "]", sep = "")}) %>%
    unname()

params_OQdiff_sex_prog <- c(mus, "female", sigmas)
```

Show chain diagnostics and posterior summaries:

```{r, fig.width = 9, fig.height = 12}
traceplot(fit_OQdiff_sex_prog, pars = params_OQdiff_sex_prog, inc_warmup = FALSE)

print(fit_OQdiff_sex_prog)

len <- length(params_OQdiff_sex_prog)
sample_plots(samples_OQdiff_sex_prog,
             params_OQdiff_sex_prog,
             cred_mass = 0.95,
             layout = matrix(c(1:(len + len %% 2)),
                             nrow = (len + len %% 2)/2,
                             ncol = 2))
```

Clean up workspace:

```{r, results = 'hide'}
cleanup("OQdiff_sex_prog")
rm(final_model, len, some_programs, mus, sigmas)
gc()
```


## Partial pooling model

This is a varying-intercept model that partially pools the programs toward the pooled mean.

Gather data to pass to Stan:

```{r}
# model name is 'OQdiff_sex_prog_hier'

data_OQdiff_sex_prog_hier <- list(y = natsap_tidy$OQdiff,
                           n = length(natsap_tidy$OQdiff),
                           sex = ifelse(natsap_tidy$sex == "MALE", 0, 1),
                           J = NROW(dose_tidy),
                           prog = natsap_tidy$new_ID)
```

Write Stan model:

```{r}
cat("
    data {
        int<lower = 1> n;                       // clients
        int<lower = 1> J;                       // programs
        vector<lower = -256, upper = 256>[n] y; // OQ diffs
        int prog[n];                            // program for each client
        vector<lower = 0, upper = 1>[n] sex;    // sex
    }
    parameters {
        real mu_a;                              // prior mean on alpha (intercepts)
        real female;                              // coefficient of sex
        real<lower = 0> sigma_a;                // prior sd on alpha (intercepts)
        vector<lower = -256, upper = 256>[J] a; // program-level intercepts
        real<lower = 0> sigma_y;                // data-level sd
    }
    model {
        // Implicit uniform priors on mu_a, sigma_a, and sigma_y
        a ~ normal(mu_a, sigma_a);
        for (i in 1:n) {
            y[i] ~ normal(a[prog[i]] + female * sex[i], sigma_y);
        }
    }
",
file = "./OQdiff_sex/OQdiff_sex_prog_hier.stan")
```

Fit Stan model:

```{r}
final_model = TRUE

if (file.exists("./OQdiff_sex/OQdiff_sex_prog_hier.rds") & final_model == TRUE) {
    stan_OQdiff_sex_prog_hier <- readRDS("./OQdiff_sex/OQdiff_sex_prog_hier.rds")
} else {
    stan_OQdiff_sex_prog_hier <- stan_model("./OQdiff_sex/OQdiff_sex_prog_hier.stan")
    saveRDS(stan_OQdiff_sex_prog_hier, file = "./OQdiff_sex/OQdiff_sex_prog_hier.rds")
}

fit_OQdiff_sex_prog_hier <- sampling(stan_OQdiff_sex_prog_hier,
                                   data = data_OQdiff_sex_prog_hier)
```

Extract and save samples:

```{r}
samples_OQdiff_sex_prog_hier <- as.data.frame(fit_OQdiff_sex_prog_hier)

write_csv(samples_OQdiff_sex_prog_hier, path = "./OQdiff_sex/OQdiff_sex_prog_hier.csv")

# Get 4 random programs
set.seed(1234)
some_programs <- sort(sample(1:data_OQdiff_sex_prog_hier$J, 4))

# Display sample sizes for selected programs.
cat("Sample sizes for programs", some_programs, ": ",
    dose_tidy[dose_tidy$new_ID %in% some_programs,"n"])

# Gather up parameter names
alphas <- some_programs %>%
    as.character() %>%
    sapply(function(x) {paste("a[", x, "]", sep = "")}) %>%
    unname()

params_OQdiff_sex_prog_hier <- c("mu_a", "sigma_a", "female",  "sigma_y", alphas)
```

Show chain diagnostics and posterior summaries:

```{r, fig.width = 9, fig.height = 10}
traceplot(fit_OQdiff_sex_prog_hier, pars = params_OQdiff_sex_prog_hier, inc_warmup = FALSE)

print(fit_OQdiff_sex_prog_hier)

len <- length(params_OQdiff_sex_prog_hier)
sample_plots(samples_OQdiff_sex_prog_hier,
             params_OQdiff_sex_prog_hier,
             cred_mass = 0.95,
             layout = matrix(c(1:(len + len %% 2)),
                             nrow = (len + len %% 2)/2,
                             ncol = 2, byrow = TRUE))
```

Clean up workspace:

```{r, results = 'hide'}
cleanup("OQdiff_sex_prog_hier")
rm(final_model, len, some_programs, alphas)
gc()
```

```{r}
beep()
```
