---
title: "NATSAP- OQ diff by RTC/WT and sex"
author: "Sean Raleigh"
date: "July 12, 2015"
output: html_document
---

This document contains the analyis of OQ difference with RTC/WT and sex as a predictor.

## Housekeeping

Load necessary libraries and special functions.

```{r, message = FALSE}
library(rstan)
library(parallel)
library(beepr)
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(stringr)
library(tidyr)
library(lme4)
```

```{r}
source("stan_utilities.R")
```

## Import data

```{r}
natsap_tidy <- read_csv("./tidydata/natsap_tidy.csv")
dose_tidy <- read_csv("./tidydata/dose_tidy.csv")
```


## No interactions

Here we model OQdiff using RTCWT and sex as predictors with no interactions.

Gather data to pass to Stan:

```{r}
# model name is 'OQdiff_RTCWT_sex'

RTCWT <- dose_tidy$RTCWT[natsap_tidy$new_ID]
RTCWT <- ifelse(RTCWT == "RTC", 0, 1)  # 0 = RTC, 1 = WT

data_OQdiff_RTCWT_sex <- list(y = natsap_tidy$OQdiff,
                              n = length(natsap_tidy$OQdiff),
                              RTCWT = RTCWT,
                              sex = ifelse(natsap_tidy$sex == "MALE", 0, 1))
```

Mathematical model:

$$ y_{i} \sim N\left( \beta_{0} + \beta_{female} sex
        + \beta_{WT} RTCWT, \sigma \right), 
        \quad 1 \leq i \leq n $$

Write Stan model:

```{r}
cat("
    data {
        int<lower = 1> n;                       // clients
        vector<lower = -256, upper = 256>[n] y; // OQ diffs
        vector<lower = 0, upper = 1>[n] RTCWT;  // RTC/WT
        vector<lower = 0, upper = 1>[n] sex;    // sex
    }
    parameters {
        real<lower = -256, upper = 256> intercept; // intercept
        real<lower = -256, upper = 256> female;    // coefficient of sex
        real<lower = -256, upper = 256> WT;        // coefficient of RTCWT
        real<lower = 0> sigma;                     // sd
    }
    model {
        // Implicit uniform priors on intercept, female, WT, and sigma
        y ~ normal(intercept + female * sex + WT * RTCWT, sigma);
    }
",
file = "./OQdiff_RTCWT_sex/OQdiff_RTCWT_sex.stan")
```

Fit Stan model:

```{r}
final_model = TRUE

if (file.exists("./OQdiff_RTCWT_sex/OQdiff_RTCWT_sex.rds") & final_model == TRUE) {
    stan_OQdiff_RTCWT_sex <- readRDS("./OQdiff_RTCWT_sex/OQdiff_RTCWT_sex.rds")
} else {
    stan_OQdiff_RTCWT_sex <- stan_model("./OQdiff_RTCWT_sex/OQdiff_RTCWT_sex.stan")
    saveRDS(stan_OQdiff_RTCWT_sex, file = "./OQdiff_RTCWT_sex/OQdiff_RTCWT_sex.rds")
}

fit_OQdiff_RTCWT_sex <- sampling(stan_OQdiff_RTCWT_sex, data = data_OQdiff_RTCWT_sex)
```

Extract and save samples:

```{r}
samples_OQdiff_RTCWT_sex <- as.data.frame(fit_OQdiff_RTCWT_sex)

write_csv(samples_OQdiff_RTCWT_sex, path = "./OQdiff_RTCWT_sex/OQdiff_RTCWT_sex.csv")

params_OQdiff_RTCWT_sex <- c("intercept", "female", "WT", "sigma")
```

Show chain diagnostics and posterior summaries:

```{r, fig.width = 9}
traceplot(fit_OQdiff_RTCWT_sex, pars = params_OQdiff_RTCWT_sex, inc_warmup = FALSE)

print(fit_OQdiff_RTCWT_sex)

len <- length(params_OQdiff_RTCWT_sex)
sample_plots(samples_OQdiff_RTCWT_sex,
             params = params_OQdiff_RTCWT_sex,
             cred_mass = 0.95,
             layout = matrix(c(1:(len + len %% 2)),
                             nrow = (len + len %% 2)/2,
                             ncol = 2))
```

Clean up workspace:

```{r, results = 'hide'}
cleanup("OQdiff_RTCWT_sex")
rm(final_model, len, RTCWT)
gc()
```


## With interaction term

Here we model OQdiff using RTCWT and sex as predictors with their interaction also.

Gather data to pass to Stan:

```{r}
# model name is 'OQdiff_RTCWT_by_sex'

RTCWT <- dose_tidy$RTCWT[natsap_tidy$new_ID]
RTCWT <- ifelse(RTCWT == "RTC", 0, 1)  # 0 = RTC, 1 = WT

data_OQdiff_RTCWT_by_sex <- list(y = natsap_tidy$OQdiff,
                                 n = length(natsap_tidy$OQdiff),
                                 RTCWT = RTCWT,
                                 sex = ifelse(natsap_tidy$sex == "MALE", 0, 1))
```

Mathematical model:

$$ y_{i} \sim N\left( \beta_{0} + \beta_{female} sex_{i} + \beta_{WT} RTCWT_{i} + 
        \beta_{WT \cdot female} (RTCWT_{i} \cdot sex_{i}), \sigma \right),
        \quad 1 \leq i \leq n $$

Write Stan model:

```{r}
cat("
    data {
        int<lower = 1> n;                       // clients
        vector<lower = -256, upper = 256>[n] y; // OQ diffs
        vector<lower = 0, upper = 1>[n] RTCWT;  // RTC/WT
        vector<lower = 0, upper = 1>[n] sex;    // sex
    }
    parameters {
        real<lower = -256, upper = 256> intercept; // intercept
        real<lower = -256, upper = 256> female;    // coefficient of sex
        real<lower = -256, upper = 256> WT;        // coefficient of RTCWT
        real<lower = -256, upper = 256> WT_female; // coefficient of interaction
        real<lower = 0> sigma;                     // sd
    }
    model {
        // Implicit uniform priors on intercept, female, WT, WT_female, and sigma
        y ~ normal(intercept + female * sex + WT * RTCWT +
            WT_female * (RTCWT .* sex), sigma);
    }
",
file = "./OQdiff_RTCWT_sex/OQdiff_RTCWT_by_sex.stan")
```

Fit Stan model:

```{r}
final_model = TRUE

if (file.exists("./OQdiff_RTCWT_sex/OQdiff_RTCWT_by_sex.rds") & final_model == TRUE) {
    stan_OQdiff_RTCWT_by_sex <- readRDS("./OQdiff_RTCWT_sex/OQdiff_RTCWT_by_sex.rds")
} else {
    stan_OQdiff_RTCWT_by_sex <- stan_model("./OQdiff_RTCWT_sex/OQdiff_RTCWT_by_sex.stan")
    saveRDS(stan_OQdiff_RTCWT_by_sex,
            file = "./OQdiff_RTCWT_sex/OQdiff_RTCWT_by_sex.rds")
}

fit_OQdiff_RTCWT_by_sex <- sampling(stan_OQdiff_RTCWT_by_sex,
                                    data = data_OQdiff_RTCWT_by_sex)
```

Extract and save samples:

```{r}
samples_OQdiff_RTCWT_by_sex <- as.data.frame(fit_OQdiff_RTCWT_by_sex)

write_csv(samples_OQdiff_RTCWT_by_sex,
          path = "./OQdiff_RTCWT_sex/OQdiff_RTCWT_by_sex.csv")

params_OQdiff_RTCWT_by_sex <- c("intercept", "female", "WT", "WT_female", "sigma")
```

Show chain diagnostics and posterior summaries:

```{r, fig.width = 9, fig.height = 10}
traceplot(fit_OQdiff_RTCWT_by_sex, pars = params_OQdiff_RTCWT_by_sex, inc_warmup = FALSE)

print(fit_OQdiff_RTCWT_by_sex)

len <- length(params_OQdiff_RTCWT_by_sex)
sample_plots(samples_OQdiff_RTCWT_by_sex,
             params = params_OQdiff_RTCWT_by_sex,
             cred_mass = 0.95,
             layout = matrix(c(1:(len + len %% 2)),
                             nrow = (len + len %% 2)/2,
                             ncol = 2, byrow = TRUE))
```

Clean up workspace:

```{r, results = 'hide'}
cleanup("OQdiff_RTCWT_by_sex")
rm(final_model, len, RTCWT)
gc()
```


## Hierarchical model

Now we consider the RTCWT by sex model with programs hierarchically modeled.

Gather data to pass to Stan:

```{r}
# model name is 'OQdiff_RTCWT_by_sex_hier'

RTCWT <- dose_tidy$RTCWT[natsap_tidy$new_ID]
RTCWT <- ifelse(RTCWT == "RTC", 0, 1)  # 0 = RTC, 1 = WT

data_OQdiff_RTCWT_by_sex_hier <- list(y = natsap_tidy$OQdiff,
                                      n = length(natsap_tidy$OQdiff),
                                      RTCWT = RTCWT,
                                      sex = ifelse(natsap_tidy$sex == "MALE", 0, 1),
                                      J = NROW(dose_tidy),
                                      prog = natsap_tidy$new_ID)
```

Mathematical model:

$$ \alpha_{j} \sim N\left( \mu_{\alpha}, \sigma_{\alpha} \right) $$
$$ y_{i} \sim ~ N\left( \alpha_{j[i]} + \beta_{female} sex_{i} + 
        \beta_{WT} RTCWT_{i} +
        \beta_{WT \cdot female}  (RTCWT_{i} \cdot sex_{i} \right), \sigma_{y}) $$

Write Stan model:

```{r}
cat("
    data {
        int<lower = 1> n;                       // clients
        vector<lower = -256, upper = 256>[n] y; // OQ diffs
        vector<lower = 0, upper = 1>[n] RTCWT;  // RTC/WT
        vector<lower = 0, upper = 1>[n] sex;    // sex
        int<lower = 1> J;                       // number of programs
        int prog[n];                            // programs
    }
    parameters {
        real<lower = -256, upper = 256> mu_a;      // program-level mean
        real<lower = 0> sigma_a;                   // program-level sd 
        vector<lower = -256, upper = 256>[J] a;    // intercept for each program
        real<lower = -256, upper = 256> female;    // coefficient of sex
        real<lower = -256, upper = 256> WT;        // coefficient of RTCWT
        real<lower = -256, upper = 256> WT_female; // coefficient of interaction
        real<lower = 0> sigma_y;                   // data-level sd
    }
    model {
        // Implicit uniform priors on intercept, female, WT, and sigma
        a ~ normal(mu_a, sigma_a);
        for(i in 1:n) {
            y[i] ~ normal(a[prog[i]] + female * sex[i] + WT * RTCWT[i] + 
                            WT_female * (RTCWT[i] * sex[i]), sigma_y);
        }
    }
",
file = "./OQdiff_RTCWT_sex/OQdiff_RTCWT_by_sex_hier.stan")
```

Fit Stan model:

```{r}
final_model = TRUE

if (file.exists("./OQdiff_RTCWT_sex/OQdiff_RTCWT_by_sex_hier.rds") & final_model == TRUE) {
    stan_OQdiff_RTCWT_by_sex_hier <- readRDS("./OQdiff_RTCWT_sex/OQdiff_RTCWT_by_sex_hier.rds")
} else {
    stan_OQdiff_RTCWT_by_sex_hier <- stan_model("./OQdiff_RTCWT_sex/OQdiff_RTCWT_by_sex_hier.stan")
    saveRDS(stan_OQdiff_RTCWT_by_sex_hier, file = "./OQdiff_RTCWT_sex/OQdiff_RTCWT_by_sex_hier.rds")
}

fit_OQdiff_RTCWT_by_sex_hier <- sampling(stan_OQdiff_RTCWT_by_sex_hier, data = data_OQdiff_RTCWT_by_sex_hier)
```

Extract and save samples:

```{r}
samples_OQdiff_RTCWT_by_sex_hier <- as.data.frame(fit_OQdiff_RTCWT_by_sex_hier)

write_csv(samples_OQdiff_RTCWT_by_sex_hier, path = "./OQdiff_RTCWT_sex/OQdiff_RTCWT_by_sex_hier.csv")

# Get 5 random programs
set.seed(1234)
some_programs <- sort(sample(1:data_OQdiff_RTCWT_by_sex_hier$J, 5))

# Display sample sizes for selected programs.
cat("Sample sizes for programs", some_programs, ": ",
    dose_tidy$n[dose_tidy$new_ID %in% some_programs])

# Gather up parameter names
alphas <- some_programs %>%
    as.character() %>%
    sapply(function(x) {paste("a[", x, "]", sep = "")}) %>%
    unname()


params_OQdiff_RTCWT_by_sex_hier <- c("mu_a", "sigma_a", 
                                     "female", "WT", "WT_female",
                                     alphas, "sigma_y")
```

Show chain diagnostics and posterior summaries:

```{r, fig.width = 9, fig.height = 10}
traceplot(fit_OQdiff_RTCWT_by_sex_hier, pars = params_OQdiff_RTCWT_by_sex_hier, inc_warmup = FALSE)

print(fit_OQdiff_RTCWT_by_sex_hier)

len <- length(params_OQdiff_RTCWT_by_sex_hier)
sample_plots(samples_OQdiff_RTCWT_by_sex_hier,
             params = params_OQdiff_RTCWT_by_sex_hier,
             cred_mass = 0.95,
             layout = matrix(c(1:12),
                             nrow = 4,
                             ncol = 3, byrow = TRUE))
```

Clean up workspace:

```{r, results = 'hide'}
cleanup("OQdiff_RTCWT_by_sex_hier")
rm(final_model, len, RTCWT, some_programs, alphas)
gc()
```

## RTCWT as a group level predictor

Rather than having RTCWT (and its interaction with sex) in the data-level model, it makes more sense to think of this as a group-level predictor. To capture the interaction between RTCWT and sex, we'll need to use a varying-intercept/varying-slope model. Unfortunately, there are programs for which we don't have data for both sexes. One way ti fix this is to include an indicator variable that is 0 for single-sex programs and 1 for coed programs. This will have the effect of removing the `sex` predictor (and therefore removing the need to estimate the slope) for programs that are not coed.

Gather data to pass to Stan:

```{r}
# model name is 'OQdiff_RTCWT_by_sex_hier2'

data_OQdiff_RTCWT_by_sex_hier2 <- list(y = natsap_tidy$OQdiff,
                                    n = length(natsap_tidy$OQdiff),
                                    RTCWT = ifelse(dose_tidy$RTCWT == "RTC", 0, 1),
                                        # 0 = RTC, 1 = WT
                                    sex = ifelse(natsap_tidy$sex == "MALE", 0, 1),
                                    J = NROW(dose_tidy),
                                    prog = natsap_tidy$new_ID,
                                    coed = ifelse(dose_tidy$coed == "Coed", 1, 0))
```

Mathematical model:

$$ \alpha_{j} \sim N\left( \gamma^{\alpha}_{0} + \gamma^{\alpha}_{1} RTCWT_{j},
        \sigma_{\alpha} \right), \quad 1 \leq j \leq J $$
$$ \beta_{j} \sim N\left( \gamma^{\beta}_{0} + \gamma^{\beta}_{1} RTCWT_{j},
        \sigma_{\beta} \right), \quad 1 \leq j \leq J $$
$$ y_{i} \sim N\left( \alpha_{j[i]} + \beta_{j[i]} * coed * sex_{i},
        \sigma_{y} \right), \quad 1 \leq i \leq n $$

Write Stan model:

```{r}
cat("
    data {
        int<lower = 1> n;                       // clients
        int<lower = 1> J;                       // number of programs
        vector<lower = -256, upper = 256>[n] y; // OQ diffs
        vector<lower = 0, upper = 1>[J] RTCWT;  // RTC/WT
        vector<lower = 0, upper = 1>[n] sex;    // sex
        int prog[n];                            // programs
        real coed[J];                            // coed or not
    }
    parameters {
        vector<lower = -256, upper = 256>[J] a;    // program-level intercepts
        real<lower = -256, upper = 256> g_a_int;   // program-level intercept
        real<lower = -256, upper = 256> g_a_slope; // coefficient of RTCWT
        real<lower = 0> sigma_a;                   // program-level sd
        vector<lower = -256, upper = 256>[J] b;    // program-level slopes
        real<lower = -256, upper = 256> g_b_int;   // coefficient of sex
        real<lower = -256, upper = 256> g_b_slope; // coefficient of interaction
        real<lower = 0> sigma_b;                   // program-level sd 
        real<lower = 0> sigma_y;                   // data-level sd
    }
    model {
        // Implicit uniform priors on g_a_int, g_a_slope, sigma_a,
        // g_b_int, g_b_slope, sigma_b,
        a ~ normal(g_a_int + g_a_slope * RTCWT, sigma_a);
        for (j in 1:J) {
            if (coed[j] == 1) {
                b[j] ~ normal(g_b_int + g_b_slope * RTCWT[j], sigma_b);
            }
        }
        for (i in 1:n) {
            y[i] ~ normal(a[prog[i]] + b[prog[i]] * coed[prog[i]] *sex[i], sigma_y);
        }
    }
",
file = "./OQdiff_RTCWT_sex/OQdiff_RTCWT_by_sex_hier2.stan")
```

Fit Stan model:

```{r}
final_model = TRUE ##### <----- CHANGE TO TRUE ONCE THE MODEL WORKS. #####

if (file.exists("./OQdiff_RTCWT_sex/OQdiff_RTCWT_by_sex_hier2.rds") & final_model == TRUE) {
    stan_OQdiff_RTCWT_by_sex_hier2 <- readRDS("./OQdiff_RTCWT_sex/OQdiff_RTCWT_by_sex_hier2.rds")
} else {
    stan_OQdiff_RTCWT_by_sex_hier2 <- stan_model("./OQdiff_RTCWT_sex/OQdiff_RTCWT_by_sex_hier2.stan")
    saveRDS(stan_OQdiff_RTCWT_by_sex_hier2, file = "./OQdiff_RTCWT_sex/OQdiff_RTCWT_by_sex_hier2.rds")
}

fit_OQdiff_RTCWT_by_sex_hier2 <- sampling(stan_OQdiff_RTCWT_by_sex_hier2, data = data_OQdiff_RTCWT_by_sex_hier2)
```

Extract and save samples:

```{r}
samples_OQdiff_RTCWT_by_sex_hier2 <- as.data.frame(fit_OQdiff_RTCWT_by_sex_hier2)

write_csv(samples_OQdiff_RTCWT_by_sex_hier2, path = "./OQdiff_RTCWT_sex/OQdiff_RTCWT_by_sex_hier2.csv")

# Get 4 random programs
set.seed(1234)
some_programs <- sort(sample(1:data_OQdiff_RTCWT_by_sex_hier2$J, 4))

# Display sample sizes for selected programs.
cat("Sample sizes for programs", some_programs, ": ",
    dose_tidy$n[dose_tidy$new_ID %in% some_programs])

# Gather up parameter names
alphas <- some_programs %>%
    as.character() %>%
    sapply(function(x) {paste("a[", x, "]", sep = "")}) %>%
    unname()

betas <- some_programs %>%
    as.character() %>%
    sapply(function(x) {paste("b[", x, "]", sep = "")}) %>%
    unname()

params_OQdiff_RTCWT_by_sex_hier2_alphas <- c("g_a_int", "g_a_slope", "sigma_a", alphas)
params_OQdiff_RTCWT_by_sex_hier2_betas <- c("g_b_int", "g_b_slope", betas)
```

Show chain diagnostics and posterior summaries:

```{r, fig.width = 9, fig.height = 10}
##### CHANGE fig.height ^^^^^ #####

traceplot(fit_OQdiff_RTCWT_by_sex_hier2, pars = params_OQdiff_RTCWT_by_sex_hier2_alphas, inc_warmup = FALSE)
traceplot(fit_OQdiff_RTCWT_by_sex_hier2, pars = params_OQdiff_RTCWT_by_sex_hier2_betas, inc_warmup = FALSE)

print(fit_OQdiff_RTCWT_by_sex_hier2)

len <- length(params_OQdiff_RTCWT_by_sex_hier2_alphas)
sample_plots(samples_OQdiff_RTCWT_by_sex_hier2,
             params = params_OQdiff_RTCWT_by_sex_hier2_alphas,
             cred_mass = 0.95,
             layout = matrix(c(1:(len + len %% 2)),
                             nrow = (len + len %% 2)/2,
                             ncol = 2, byrow = TRUE))

len <- length(params_OQdiff_RTCWT_by_sex_hier2_betas)
sample_plots(samples_OQdiff_RTCWT_by_sex_hier2,
             params = params_OQdiff_RTCWT_by_sex_hier2_betas,
             cred_mass = 0.95,
             layout = matrix(c(1:(len + len %% 2)),
                             nrow = (len + len %% 2)/2,
                             ncol = 2, byrow = TRUE))
```

Clean up workspace:

```{r, results = 'hide'}
cleanup("OQdiff_RTCWT_by_sex_hier2")
rm(final_model, len, some_programs, alphas, betas, params_OQdiff_RTCWT_by_sex_hier2_alphas, params_OQdiff_RTCWT_by_sex_hier2_betas) 
gc()
```



```{r}
beep()
```
