---
title: "NATSAP- OQ diff"
author: "Sean Raleigh"
date: "August 19, 2015"
output: html_document
---

This document contains the simplest analysis for OQ differences with no predictors. This is just a test of the same thing in `OQdiff` except with scaled variables.

## Housekeeping

Load necessary libraries and special functions.

```{r, message = FALSE}
library(rstan)
library(parallel)
library(beepr)
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(stringr)
library(tidyr)
library(lme4)
```

```{r}
source("stan_utilities.R")
```


## Import data

```{r}
natsap_tidy <- read.csv("./tidydata/natsap_tidy.csv")
dose_tidy <- read.csv("./tidydata/dose_tidy.csv")
dose_tidy_z <- read.csv("./tidydata/dose_tidy_z.csv")
```


## Partial pooling model

This is a varying-intecept hierarchical model. This allows partial pooling and will cause the programs with small sample sizes to shrink to the complete pooling mean.

Gather data to pass to Stan:

```{r}
# Model name is 'OQdiff_z_prog_hier'

data_OQdiff_z_prog_hier <- list(y = natsap_tidy$OQdiff,
                            n = length(natsap_tidy$OQdiff),
                            J = NROW(dose_tidy_z),
                            prog = natsap_tidy$new_ID)
```

Mathematical model:

$$ \alpha_{j} \sim N\left(\mu_{\alpha}, \sigma_{\alpha}\right), \quad 1 \leq j \leq J $$
$$ y_{i} \sim N\left(\alpha_{j[i]}, \sigma_{y}\right), \quad 1 \leq i \leq n $$

Write Stan model:

```{r}
cat("
    data {
        int<lower = 1> n;                       // clients
        int<lower = 1> J;                       // programs
        vector<lower = -256, upper = 256>[n] y; // OQ diffs
        int prog[n];                            // program for each client
    }
    parameters {
        real<lower = -256, upper = 256> mu_a;   // prior mean on alpha (intercepts)
        real<lower = 0> sigma_a;                // prior sd on alpha (intercepts)
        vector<lower = -256, upper = 256>[J] a; // program-level intercepts
        real<lower = 0> sigma_y;                // data-level sd
    }
    model {
        // Implicit uniform priors on mu_a, sigma_a, and sigma_y
        a ~ normal(mu_a, sigma_a);
        for (i in 1:n) {
            y[i] ~ normal(a[prog[i]], sigma_y);
        }
    }
",
file = "./OQdiff_z/OQdiff_z_prog_hier.stan")
```

Fit Stan model:

```{r}
final_model = TRUE

if (file.exists("./OQdiff_z/OQdiff_z_prog_hier.rds") & final_model == TRUE) {
    stan_OQdiff_z_prog_hier <- readRDS("./OQdiff_z/OQdiff_z_prog_hier.rds")
} else {
    stan_OQdiff_z_prog_hier <- stan_model("./OQdiff_z/OQdiff_z_prog_hier.stan")
    saveRDS(stan_OQdiff_z_prog_hier, file = "./OQdiff_z/OQdiff_z_prog_hier.rds")
}

fit_OQdiff_z_prog_hier <- sampling(stan_OQdiff_z_prog_hier, data = data_OQdiff_z_prog_hier)
```

Extract and save samples:

```{r}
samples_OQdiff_z_prog_hier <- as.data.frame(fit_OQdiff_z_prog_hier)

write_csv(samples_OQdiff_z_prog_hier, path = "./OQdiff_z/OQdiff_z_prog_hier.csv")

# Get 5 random programs
set.seed(1234)
some_programs <- sort(sample(1:data_OQdiff_z_prog_hier$J, 5))

# Display sample sizes for selected programs.
cat("Sample sizes for programs", some_programs, ": ",
    dose_tidy$n[dose_tidy$new_ID %in% some_programs])

# Gather up parameter names
alphas <- some_programs %>%
    as.character() %>%
    sapply(function(x) {paste("a[", x, "]", sep = "")}) %>%
    unname()

params_OQdiff_z_prog_hier <- c("mu_a", "sigma_a", alphas, "sigma_y")
```

Show chain diagnostics and posterior summaries.

```{r, fig.width = 9, fig.height = 10}
traceplot(fit_OQdiff_z_prog_hier, pars = params_OQdiff_z_prog_hier, inc_warmup = FALSE)

print(fit_OQdiff_z_prog_hier)

len <- length(params_OQdiff_z_prog_hier)
sample_plots(samples_OQdiff_z_prog_hier,
             params = params_OQdiff_z_prog_hier,
             cred_mass = 0.95,
             layout = matrix(c(1:(len + len %% 2)),
                             nrow = (len + len %% 2)/2,
                             ncol = 2, byrow = TRUE))
```

Clean up workspace.

```{r, results = 'hide'}
cleanup("OQdiff_z_prog_hier")
rm(final_model, len, some_programs, alphas)
gc()
```


```{r}
beep()
```
