---
title: "NATSAP- OQ diff by RTC/WT and sex, plus dose predictors"
author: "Sean Raleigh"
date: "August 19, 2015"
output: html_document
---

This document contains the analyis of OQ difference with RTC/WT and sex (and their interaction) as predictors, along with other group-level dose predictors. This will still be a variable-intercept model as a variable-slope model continues to elude me.

## Housekeeping

Load necessary libraries and special functions.

```{r, message = FALSE}
library(rstan)
library(parallel)
library(beepr)
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(stringr)
library(tidyr)
library(lme4)
```

```{r}
source("stan_utilities.R")
```

## Import data

```{r}
natsap_tidy <- read_csv("./tidydata/natsap_tidy.csv")
dose_tidy_z <- read_csv("./tidydata/dose_tidy_z.csv")
```


## Try everything related to dose

Here we model OQdiff using RTCWT and sex along with all other predictors relating to dose.

Gather data to pass to Stan:

```{r}
# model name is 'OQdiff_RTCWT_by_sex_dose'

RTCWT <- dose_tidy_z$RTCWT[natsap_tidy$new_ID]
RTCWT <- ifelse(RTCWT == "RTC", 0, 1)  # 0 = RTC, 1 = WT

data_OQdiff_RTCWT_by_sex_dose <- list(y = natsap_tidy$OQdiff,
                                           n = length(natsap_tidy$OQdiff),
                                           J = NROW(dose_tidy_z),
                                           prog = natsap_tidy$new_ID,
                                           RTCWT = RTCWT,
                                           sex = ifelse(natsap_tidy$sex == "MALE", 0, 1),
                                           IT = dose_tidy_z$IT,
                                           GT = dose_tidy_z$GT,
                                           RFT = dose_tidy_z$RFT,
                                           IPFT = dose_tidy_z$IPFT,
                                           PA = dose_tidy_z$PA,
                                           school = dose_tidy_z$school)
```

Mathematical model:

$$ \alpha_{j} \sim N\left( \mu_{\alpha} + \gamma_{IT} IT + \dots + \gamma_{school} school, \sigma_{\alpha} \right) $$
$$ y_{i} \sim ~ N\left( \alpha_{j[i]} + \beta_{female} sex_{i} + 
        \beta_{WT} RTCWT_{i} +
        \beta_{WT \cdot female}  (RTCWT_{i} \cdot sex_{i} \right), \sigma_{y}) $$

Write Stan model:

```{r}
cat("
    data {
        int<lower = 1> n;                       // clients
        vector<lower = -256, upper = 256>[n] y; // OQ diffs
        vector<lower = 0, upper = 1>[n] RTCWT;  // RTC/WT
        vector<lower = 0, upper = 1>[n] sex;    // sex
        int<lower = 1> J;                       // number of programs
        int prog[n];                            // programs
        vector[J] IT;                           // Ind. therapy
        vector[J] GT;                           // Group therapy
        vector[J] RFT;                          // Remote family therapy
        vector[J] IPFT;                         // In-person family therapy
        vector[J] PA;                           // Physical activities
        vector[J] school;                       // School
    }
    parameters {
        real mu_a;                // program-level mean
        real<lower = 0> sigma_a;  // program-level sd 
        vector[J] a;              // intercept for each program
        real g_IT;                // Coef. of Ind. therapy
        real g_GT;                // Coef. of Group therapy
        real g_RFT;               // Coef. of Remote family therapy
        real g_IPFT;              // Coef. of In-person family therapy
        real g_PA;                // Coef. of Physical activities
        real g_school;            // Coef. of School
        real female;              // coefficient of sex
        real WT;                  // coefficient of RTCWT
        real WT_female;           // coefficient of interaction
        real<lower = 0> sigma_y;  // data-level sd
    }
    model {
        // Implicit uniform priors on everrything
        a ~ normal(mu_a + g_IT * IT + g_GT * GT +
                    g_RFT * RFT + g_IPFT * IPFT +
                    g_PA * PA + g_school * school, sigma_a);
        for(i in 1:n) {
            y[i] ~ normal(a[prog[i]] + female * sex[i] + WT * RTCWT[i] + 
                            WT_female * (RTCWT[i] * sex[i]), sigma_y);
        }
    }
",
file = "./OQdiff_RTCWT_by_sex_dose/OQdiff_RTCWT_by_sex_dose.stan")
```

Fit Stan model:

```{r}
final_model = TRUE

if (file.exists("./OQdiff_RTCWT_by_sex_dose/OQdiff_RTCWT_by_sex_dose.rds") & final_model == TRUE) {
    stan_OQdiff_RTCWT_by_sex_dose <-            readRDS("./OQdiff_RTCWT_by_sex_dose/OQdiff_RTCWT_by_sex_dose.rds")
} else {
    stan_OQdiff_RTCWT_by_sex_dose <- stan_model("./OQdiff_RTCWT_by_sex_dose/OQdiff_RTCWT_by_sex_dose.stan")
    saveRDS(stan_OQdiff_RTCWT_by_sex_dose, file = "./OQdiff_RTCWT_by_sex_dose/OQdiff_RTCWT_by_sex_dose.rds")
}

fit_OQdiff_RTCWT_by_sex_dose <- sampling(stan_OQdiff_RTCWT_by_sex_dose, data = data_OQdiff_RTCWT_by_sex_dose)
```

Extract and save samples:

```{r}
samples_OQdiff_RTCWT_by_sex_dose <- as.data.frame(fit_OQdiff_RTCWT_by_sex_dose)

write_csv(samples_OQdiff_RTCWT_by_sex_dose, path = "./OQdiff_RTCWT_by_sex_dose/OQdiff_RTCWT_by_sex_dose.csv")

# Get 5 random programs
set.seed(1234)
some_programs <- sort(sample(1:data_OQdiff_RTCWT_by_sex_dose$J, 5))

# Display sample sizes for selected programs.
cat("Sample sizes for programs", some_programs, ": ",
    dose_tidy_z$n[dose_tidy_z$new_ID %in% some_programs])

# Gather up parameter names
alphas <- some_programs %>%
    as.character() %>%
    sapply(function(x) {paste("a[", x, "]", sep = "")}) %>%
    unname()


params_OQdiff_RTCWT_by_sex_dose <- c("mu_a", "sigma_a",
                                          # alphas,
                                          "female", "WT", "WT_female",
                                          "g_IT", "g_GT",
                                          "g_RFT", "g_IPFT",
                                          "g_PA", "g_school",
                                          "sigma_y")
```

Show chain diagnostics and posterior summaries:

```{r, fig.width = 9, fig.height = 10}
traceplot(fit_OQdiff_RTCWT_by_sex_dose, pars = params_OQdiff_RTCWT_by_sex_dose, inc_warmup = FALSE)

print(fit_OQdiff_RTCWT_by_sex_dose)

len <- length(params_OQdiff_RTCWT_by_sex_dose)
sample_plots(samples_OQdiff_RTCWT_by_sex_dose,
             params = params_OQdiff_RTCWT_by_sex_dose,
             cred_mass = 0.95,
             layout = matrix(c(1:12),
                             nrow = 4,
                             ncol = 3, byrow = TRUE))
```

Clean up workspace:

```{r, results = 'hide'}
cleanup("OQdiff_RTCWT_by_sex_dose")
rm(final_model, len, RTCWT, some_programs, alphas)
gc()
```


```{r}
beep()
```
