---
title: "NATSAP- OQ diff"
author: "Sean Raleigh"
date: "July 10, 2015"
output: html_document
---

This document contains the simplest analysis for OQ differences with no predictors.

## Housekeeping

Load necessary libraries and special functions.

```{r, message = FALSE}
library(rstan)
library(parallel)
library(beepr)
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(stringr)
library(tidyr)
library(lme4)
```

```{r}
source("stan_utilities.R")
```


## Import data

```{r}
natsap_tidy <- read_csv("./tidydata/natsap_tidy_1.csv")
dose_tidy <- read.csv("./tidydata/dose_tidy_1.csv")
```


## Complete pooling model

Consider `OQdiff` alone. (All programs are averaged together.)

Gather data to pass to Stan:

```{r}
# Model name is 'OQdiff'

data_OQdiff <- list(y = natsap_tidy$OQdiff,
                  n = length(natsap_tidy$OQdiff))
```

Write Stan model:

```{r}
cat("
    data {
        int<lower = 1> n;                       // clients
        vector<lower = -256, upper = 256>[n] y; // OQ diffs
    }
    parameters {
        real<lower = -256, upper = 256> mu;
        real<lower = 0> sigma;
    }
    model {
        // Implicit uniform priors on mu and sigma
        y ~ normal(mu, sigma);
    }
",
file = "./OQdiff/OQdiff.stan")
```

Fit Stan model:

```{r}
final_model = TRUE

if (file.exists("./OQdiff/OQdiff.rds") & final_model == TRUE) {
    stan_OQdiff <- readRDS("./OQdiff/OQdiff.rds")
} else {
    stan_OQdiff <- stan_model("./OQdiff/OQdiff.stan")
    saveRDS(stan_OQdiff, file = "./OQdiff/OQdiff.rds")
}

fit_OQdiff <- sampling(stan_OQdiff, data = data_OQdiff)
```

Extract and save samples.

```{r}
samples_OQdiff <- as.data.frame(fit_OQdiff)

write_csv(samples_OQdiff, path = "./OQdiff/OQdiff.csv")

params_OQdiff <- c("mu", "sigma")
```

Show chain diagnostics and posterior summaries.

```{r, fig.width = 9}
traceplot(fit_OQdiff, pars = params_OQdiff, inc_warmup = FALSE)

print(fit_OQdiff)

len <- length(params_OQdiff)
sample_plots(samples_OQdiff,
             params = params_OQdiff,
             cred_mass = 0.95,
             layout = matrix(c(1:(len + len %% 2)),
                             nrow = (len + len %% 2)/2,
                             ncol = 2))
```

Clean up workspace.

```{r, results = 'hide'}
cleanup("OQdiff")
rm(final_model, len)
gc()
```


## No pooling model

Now consider `OQdiff` by program. (In other words, each program is modeled as completely separate from other programs.)

Gather data to pass to Stan:

```{r}
# Model name is 'OQdiff_prog'

data_OQdiff_prog <- list(y = natsap_tidy$OQdiff,
                       n = length(natsap_tidy$OQdiff),
                       J = NROW(dose_tidy),
                       prog = natsap_tidy$new_ID)
```

Write Stan model:

```{r}
cat("
    data {
        int<lower = 1> n;                       // clients
        int<lower = 1> J;                       // programs
        vector<lower = -256, upper = 256>[n] y; // OQ diffs
        int prog[n];                            // program for each client
    }
    parameters {
        vector<lower = -256, upper = 256>[J] mu;
        vector<lower = 0>[J] sigma;
    }
    model {
        // Implicit uniform priors on mus and sigmas
        for (i in 1:n) {
            y[i] ~ normal(mu[prog[i]], sigma[prog[i]]);
        }
    }
",
file = "./OQdiff/OQdiff_prog.stan")
```

Fit Stan model:

```{r}
final_model = TRUE

if (file.exists("./OQdiff/OQdiff_prog.rds") & final_model == TRUE) {
    stan_OQdiff_prog <- readRDS("./OQdiff/OQdiff_prog.rds")
} else {
    stan_OQdiff_prog <- stan_model("./OQdiff/OQdiff_prog.stan")
    saveRDS(stan_OQdiff_prog, file = "./OQdiff/OQdiff_prog.rds")
}

fit_OQdiff_prog <- sampling(stan_OQdiff_prog, data = data_OQdiff_prog)
```

Extract and save samples:

```{r}
samples_OQdiff_prog <- as.data.frame(fit_OQdiff_prog)

write_csv(samples_OQdiff_prog, path = "./OQdiff/OQdiff_prog.csv")

# Get 4 random programs
set.seed(1234)
some_programs <- sort(sample(1:data_OQdiff_prog$J, 4))

# Display sample sizes for selected programs.
cat("Sample sizes for programs", some_programs, ": ",
    dose_tidy[dose_tidy$new_ID %in% some_programs,"n"])

# Gather up parameter names
mus <- some_programs %>%
    as.character() %>%
    sapply(function(x) {paste("mu[", x, "]", sep = "")}) %>%
    unname()
sigmas <- some_programs %>%
    as.character() %>%
    sapply(function(x) {paste("sigma[", x, "]", sep = "")}) %>%
    unname()

params_OQdiff_prog <- c(mus, sigmas)
```

Show chain diagnostics and posterior summaries.

```{r, fig.width = 9, fig.height = 10}
traceplot(fit_OQdiff_prog, pars = params_OQdiff_prog, inc_warmup = FALSE)

print(fit_OQdiff_prog)

len <- length(params_OQdiff_prog)
sample_plots(samples_OQdiff_prog,
             params = params_OQdiff_prog,
             cred_mass = 0.95,
             layout = matrix(c(1:(len + len %% 2)),
                             nrow = (len + len %% 2)/2,
                             ncol = 2))
```

Clean up workspace

```{r, results = 'hide'}
cleanup("OQdiff_prog")
rm(final_model, len, some_programs, mus, sigmas)
gc()
```


## Partial pooling model

This is a varying-intecept hierarchical model. This allows partial pooling and will cause the programs with small sample sizes to shrink to the complete pooling mean.

Gather data to pass to Stan:

```{r}
# Model name is 'OQdiff_prog_hier'

data_OQdiff_prog_hier <- list(y = natsap_tidy$OQdiff,
                            n = length(natsap_tidy$OQdiff),
                            J = NROW(dose_tidy),
                            prog = natsap_tidy$new_ID)
```

Write Stan model:

```{r}
cat("
    data {
        int<lower = 1> n;                       // clients
        int<lower = 1> J;                       // programs
        vector<lower = -256, upper = 256>[n] y; // OQ diffs
        int prog[n];                            // program for each client
    }
    parameters {
        real mu_a;                              // prior mean on alpha (intercepts)
        real<lower = 0> sigma_a;                // prior sd on alpha (intercepts)
        vector<lower = -256, upper = 256>[J] a; // program-level intercepts
        real<lower = 0> sigma_y;                // data-level sd
    }
    model {
        // Implicit uniform priors on mu_a, sigma_a, and sigma_y
        a ~ normal(mu_a, sigma_a);
        for (i in 1:n) {
            y[i] ~ normal(a[prog[i]], sigma_y);
        }
    }
",
file = "./OQdiff/OQdiff_prog_hier.stan")
```

Fit Stan model:

```{r}
final_model = TRUE

if (file.exists("./OQdiff/OQdiff_prog_hier.rds") & final_model == TRUE) {
    stan_OQdiff_prog_hier <- readRDS("./OQdiff/OQdiff_prog_hier.rds")
} else {
    stan_OQdiff_prog_hier <- stan_model("./OQdiff/OQdiff_prog_hier.stan")
    saveRDS(stan_OQdiff_prog_hier, file = "./OQdiff/OQdiff_prog_hier.rds")
}

fit_OQdiff_prog_hier <- sampling(stan_OQdiff_prog_hier, data = data_OQdiff_prog_hier)
```

Extract and save samples:

```{r}
samples_OQdiff_prog_hier <- as.data.frame(fit_OQdiff_prog_hier)

write_csv(samples_OQdiff_prog_hier, path = "./OQdiff/OQdiff_prog_hier.csv")

# Get 5 random programs
set.seed(1234)
some_programs <- sort(sample(1:data_OQdiff_prog_hier$J, 5))

# Display sample sizes for selected programs.
cat("Sample sizes for programs", some_programs, ": ",
    dose_tidy[dose_tidy$new_ID %in% some_programs,"n"])

# Gather up parameter names
alphas <- some_programs %>%
    as.character() %>%
    sapply(function(x) {paste("a[", x, "]", sep = "")}) %>%
    unname()

params_OQdiff_prog_hier <- c("mu_a", "sigma_a", alphas, "sigma_y")
```

Show chain diagnostics and posterior summaries.

```{r, fig.width = 9, fig.height = 10}
traceplot(fit_OQdiff_prog_hier, pars = params_OQdiff_prog_hier, inc_warmup = FALSE)

print(fit_OQdiff_prog_hier)

len <- length(params_OQdiff_prog_hier)
sample_plots(samples_OQdiff_prog_hier,
             params = params_OQdiff_prog_hier,
             cred_mass = 0.95,
             layout = matrix(c(1:(len + len %% 2)),
                             nrow = (len + len %% 2)/2,
                             ncol = 2, byrow = TRUE))
```

Clean up workspace.

```{r, results = 'hide'}
cleanup("OQdiff_prog_hier")
rm(final_model, len, some_programs, alphas)
gc()
```

## Discussion

One can see the effect of partial pooling in the graph below.

```{r, fig.width = 9}
## Import OQdiff.csv, OQdiff_prog.csv, and OQdiff_prog_hier.csv

samples_OQdiff <- read_csv("./OQdiff/OQdiff.csv")
samples_OQdiff_prog <- read_csv("./OQdiff/OQdiff_prog.csv")
samples_OQdiff_prog_hier <- read_csv("./OQdiff/OQdiff_prog_hier.csv")


## Gather up all the parameters of interest into a single data frame

plot_stats_no_pooling <- params_per_program(samples_OQdiff_prog) %>%
    select(Program, Median = mu_median, q025 = mu_q025, q975 = mu_q975) %>%
    mutate(Model = "No pooling")
plot_stats_partial_pooling <- params_per_program(samples_OQdiff_prog_hier) %>%
    select(Program, Median = a_median, q025 = a_q025, q975 = a_q975) %>%
    mutate(Model = "Partial pooling")
plot_stats <- rbind(plot_stats_no_pooling, plot_stats_partial_pooling)


## Plot complete pooling vs no pooling vs partial pooling

ggplot(plot_stats) +
    geom_pointrange(aes(x = Program,
                        y = Median,
                        ymin = q025,
                        ymax = q975,
                        color = Model),
                   position = position_dodge(width = 0.5)) +
    geom_hline(aes(color = "#619CFF"),
               color = "#619CFF",
               yintercept = median(samples_OQdiff$mu)) + 
    scale_color_manual(values = c("No pooling" = "#F8766D",
                                  "Partial pooling" = "#00BA38",
                                  "Complete pooling" = "#619CFF"),
                       limits = c("No pooling",
                                  "Partial pooling",
                                  "Complete pooling")) +
    ylab("OQ diff (with 95% credible interval)") +
    theme(title = element_text(size = rel(1.5)))
```

The results agree with `lmer`.

```{r}
test_lmer <- lmer(OQdiff ~ 1 + (1 | new_ID), data = natsap_tidy)
fixef(test_lmer)
ranef(test_lmer)
```

Clean up workspace

```{r, results = 'hide'}
rm(samples_OQdiff, samples_OQdiff_prog, samples_OQdiff_prog_hier, test_lmer, plot_stats, plot_stats_no_pooling, plot_stats_partial_pooling)
gc()
```

```{r}
beep()
```
