---
title: "NATSAP- Presentation graphics"
author: "Sean Raleigh"
date: "August 26, 2015"
output: html_document
---

In this document, we prepare graphics to illustrate the model 

## Housekeeping

Load necessary libraries and special functions.

```{r, message = FALSE}
library(beepr)
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(tidyr)
library(grid)
```

## Import data

```{r}
natsap_tidy <- read.csv("./tidydata/natsap_tidy.csv")
dose_tidy <- read.csv("./tidydata/dose_tidy.csv")
dose_tidy_z <- read.csv("./tidydata/dose_tidy_z.csv")
OQdiff_RTCWT_by_sex_dose <- read.csv("./OQdiff_RTCWT_by_sex_dose/OQdiff_RTCWT_by_sex_dose.csv")
OQdiff_RTCWT_by_sex_freq <- read.csv("./OQdiff_RTCWT_by_sex_freq/OQdiff_RTCWT_by_sex_freq.csv")
```

## Compile necessary summaries

```{r}
# multi.sapply function from 
# http://www.r-bloggers.com/applying-multiple-functions-to-data-frame/
multi.sapply <- function(...) {
      arglist <- match.call(expand.dots = FALSE)$...
      var.names <- sapply(arglist, deparse)
      has.name <- (names(arglist) != "")
      var.names[has.name] <- names(arglist)[has.name]
      arglist <- lapply(arglist, eval.parent, n = 2)
      x <- arglist[[1]]
      arglist[[1]] <- NULL
      result <- sapply(arglist, function (FUN, x) sapply(x, FUN), x)
      colnames(result) <- var.names[-1]
      return(result)
}

# for dose

dose_summary <- OQdiff_RTCWT_by_sex_dose %>%
    select(-starts_with("a."), -starts_with("lp"), -starts_with("sigma")) %>%
    multi.sapply(q025 = function(x) {unname(quantile(x, 0.025))},
                 med = function(x) {median(x)},
                 q975 = function(x) {unname(quantile(x, 0.975))}) %>%
    as.data.frame() %>%
    cbind(vars = rownames(.))
rownames(dose_summary) <- NULL

## Set up new variable names and order
graph_order_dose <- data.frame(
    name = factor(c("Main Effect", "WT", "female", "WT * female",
             "IT", "GT", "RFT", "IPFT", "PA", "School"),
             levels = c("Main Effect", "WT", "female", "WT * female",
             "IT", "GT", "RFT", "IPFT", "PA", "School"),
             ordered = TRUE),
    vars = c("mu_a", "WT", "female", "WT_female",
             "g_IT", "g_GT", "g_RFT", "g_IPFT", "g_PA", "g_school")
)

dose_summary <- dose_summary %>%
    inner_join(graph_order_dose, by = "vars") %>%
    arrange(name)

# for frequency

freq_summary <- OQdiff_RTCWT_by_sex_freq %>%
    select(-starts_with("a."), -starts_with("lp"), -starts_with("sigma")) %>%
    multi.sapply(q025 = function(x) {unname(quantile(x, 0.025))},
                 med = function(x) {median(x)},
                 q975 = function(x) {unname(quantile(x, 0.975))}) %>%
    as.data.frame() %>%
    cbind(vars = rownames(.))
rownames(freq_summary) <- NULL

## Set up new variable names and order
graph_order_freq <- data.frame(
    name = factor(c("Main Effect", "WT", "female", "WT * female",
             "IT", "GT", "RFT", "IPFT"),
             levels = c("Main Effect", "WT", "female", "WT * female",
             "IT", "GT", "RFT", "IPFT"),
             ordered = TRUE),
    vars = c("mu_a", "WT", "female", "WT_female",
             "g_IT_freq", "g_GT_freq", "g_RFT_freq", "g_IPFT_freq")
)

freq_summary <- freq_summary %>%
    inner_join(graph_order_freq, by = "vars") %>%
    arrange(name)

```

## Set up presentation theme

```{r}
theme_pres <- theme(
    text = element_text(size = rel(8)),
    legend.title = element_text(size = rel(6)),
    legend.key.size = unit(1, "in"),
    legend.text = element_text(size = rel(4))
)
```

## Graph for dose

```{r}
dose_graph <- ggplot(dose_summary, aes(x = name, y = med, ymin = q025, ymax = q975)) +
    geom_pointrange(size = rel(2)) +
    geom_hline(y = 0, color = "blue") +
    xlab("Dose Variables") +
    ylab("OQ difference") +
    theme_pres +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
dose_graph
```

## Graph for frequency

```{r}
freq_graph <- ggplot(freq_summary, aes(x = name, y = med, ymin = q025, ymax = q975)) +
    geom_pointrange(size = rel(2)) +
    geom_hline(y = 0, color = "blue") +
    xlab("Frequency Variables") +
    ylab("OQ difference") +
    theme_pres +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
freq_graph
```


## Investigate RTF_freq

The frequency of remote family therapy seems quite significant, so here we investigate. First, add a column to `dose_tidy` to get the median OQdiff for each program.

```{r}
OQdiff_prog <- natsap_tidy %>%
    group_by(ID) %>%
    summarize(median_OQdiff = median(OQdiff) * 1.0)

RFT_check <- dose_tidy %>%
    inner_join(OQdiff_prog, by = "ID")
```

Next, we plot `RTF_freq` against the median OQdiff.

```{r}
ggplot(RFT_check, aes(x = RFT_freq, y = median_OQdiff, size = n)) +
    geom_point() +
    scale_size_continuous(range = c(5,20)) +
    xlab("RFT frequency (sessions/week)") +
    ylab("OQ difference") +
    theme_pres +
    theme(legend.position="none")
```

## RTC vs WT

```{r}
OQ_explore <- natsap_tidy %>%
    inner_join(dose_tidy, by = "new_ID")

OQ_summary <- OQ_explore %>%
    group_by(RTCWT, sex) %>%
    summarize(`Median OQ_A` = median(OQ_A)*1.0,
              `Median OQ_D` = median(OQ_D)*1.0,
              `Median OQdiff` = median(OQdiff)*1.0,
              n())
OQ_summary

OQ_stats <- OQ_explore %>%
    gather(AD, OQ, OQ_A, OQ_D)
levels(OQ_stats$AD)[levels(OQ_stats$AD) == "OQ_A"] <- "Admission OQ"
levels(OQ_stats$AD)[levels(OQ_stats$AD) == "OQ_D"] <- "Discharge OQ"

plot_OQ_AD <- ggplot(OQ_stats, aes(x = RTCWT, y = OQ, fill = sex)) +
    geom_boxplot() +
    facet_grid(. ~ AD) +
    xlab("") +
    scale_fill_discrete(name = "Sex") +
    theme_pres
plot_OQ_AD

plot_OQ_diff <- ggplot(OQ_stats, aes(x = RTCWT, y = OQdiff, fill = sex)) +
    geom_boxplot() +
    xlab("") +
    ylab("OQ difference") +
    scale_fill_discrete(name = "Sex") +
    theme_pres
plot_OQ_diff
```
