---
title: "Reproducing 2014 results in Stan- Part 2"
author: "Sean Raleigh"
date: "July 10, 2015"
output: html_document
---

To get Stan working, we reproduce the results from the 2014 report, except we convert everything from `lme4` to Stan. (This is Part 2: the file is split up to keep compilation managaeable.)

This part contains the hierarchical partial pooling model of diff by program.

## Housekeeping

Load necessary libraries and special functions.

```{r, message = FALSE}
library(rstan)
library(parallel)
library(beepr)
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(stringr)
library(tidyr)
library(lme4)
```

```{r}
source("stan_utilities.R")
```

## Import data

The data sets were cleaned and tidied in Part 1, so all we have to do here is import them into our workspace.

```{r}
natsap_tidy <- read_csv("natsap_tidy_1.csv")
dose_tidy <- read.csv("dose_tidy_1.csv")
```

## Hierarchical models

### Partial pooling model, diff by program

Gather data to pass to Stan:

```{r}
# Model name is 'diff_prog_hier'

data_diff_prog_hier <- list(y = natsap_tidy$diff,
                            n = length(natsap_tidy$diff),
                            J = NROW(dose_tidy),
                            prog = natsap_tidy$new_ID)
```

Write Stan model:

```{r}
cat("
    data {
        int<lower = 1> n;                       // clients
        int<lower = 1> J;                       // programs
        vector<lower = -256, upper = 256>[n] y; // OQ diffs
        int prog[n];                            // program for each client
    }
    parameters {
        real mu_a;                              // prior mean on alpha (intercepts)
        real<lower = 0> sigma_a;                // prior sd on alpha (intercepts)
        vector<lower = -256, upper = 256>[J] a; // program-level intercepts
        real<lower = 0> sigma_y;                // data-level sd
    }
    model {
        // Implicit uniform priors on mu_a, sigma_a, and sigma_y
        a ~ normal(mu_a, sigma_a);
        for (i in 1:n) {
            y[i] ~ normal(a[prog[i]], sigma_y);
        }
    }
",
file = "diff_prog_hier.stan")
```

Fit Stan model:

```{r}
final_model = TRUE

if (file.exists("diff_prog_hier.rds") & final_model == TRUE) {
    stan_diff_prog_hier <- readRDS("diff_prog_hier.rds")
} else {
    stan_diff_prog_hier <- stan_model("diff_prog_hier.stan")
    saveRDS(stan_diff_prog_hier, file = "diff_prog_hier.rds")
}

fit_diff_prog_hier <- sampling(stan_diff_prog_hier, data = data_diff_prog_hier)
```

Extract and save samples:

```{r}
samples_diff_prog_hier <- as.data.frame(fit_diff_prog_hier)

write_csv(samples_diff_prog_hier, path = "./diff_prog_hier.csv")

# Get 5 random programs
set.seed(5678)
some_programs <- sort(sample(1:data_diff_prog_hier$J, 5))

# Display sample sizes for selected programs.
cat("Sample sizes for programs", some_programs, ": ",
    dose_tidy[dose_tidy$new_ID %in% some_programs,"n"])

# Gather up parameter names
alphas <- some_programs %>%
    as.character() %>%
    sapply(function(x) {paste("a[", x, "]", sep = "")}) %>%
    unname()

params_diff_prog_hier <- c("mu_a", "sigma_a", alphas, "sigma_y")
```

Show chain diagnostics and posterior summaries.

```{r, fig.width = 9, fig.height = 10}
traceplot(fit_diff_prog_hier, pars = params_diff_prog_hier, inc_warmup = FALSE)

print(fit_diff_prog_hier)

sample_plots(samples_diff_prog_hier,
             params_diff_prog_hier,
             cred_mass = 0.95,
             layout = matrix(c(1:length(params_diff_prog_hier)),
                             nrow = length(params_diff_prog_hier)/2,
                             ncol = 2, byrow = TRUE))
```


## Discussion

One can see the effect of partial pooling in the graph below.

```{r, fig.width = 9}
## Import samples_diff.csv and samples_diff_prog.csv

samples_diff <- read_csv("diff.csv")
samples_diff_prog <- read_csv("diff_prog.csv")

## Gather up all the parameters of interest into a single data frame

plot_stats_no_pooling <- params_per_program(samples_diff_prog) %>%
    select(Program, Median = mu_median, q05 = mu_q05, q95 = mu_q95) %>%
    mutate(Model = "No pooling")
plot_stats_partial_pooling <- params_per_program(samples_diff_prog_hier) %>%
    select(Program, Median = a_median, q05 = a_q05, q95 = a_q95) %>%
    mutate(Model = "Partial pooling")
plot_stats <- rbind(plot_stats_no_pooling, plot_stats_partial_pooling)

## Plot complete pooling vs no pooling vs partial pooling

ggplot(plot_stats) +
    geom_pointrange(aes(x = Program,
                        y = Median,
                        ymin = q05,
                        ymax = q95,
                        color = Model),
                   position = position_dodge(width = 0.5)) +
    geom_hline(aes(color = "#619CFF"),
               color = "#619CFF",
               yintercept = median(samples_diff$mu)) + 
    scale_color_manual(values = c("No pooling" = "#F8766D",
                                  "Partial pooling" = "#00BA38",
                                  "Complete pooling" = "#619CFF"),
                       limits = c("No pooling",
                                  "Partial pooling",
                                  "Complete pooling")) +
    ylab("OQ diff (with 95% credible interval)") +
    theme(title = element_text(size = rel(1.5)))
    
```

The results agree with `lmer`.

```{r}
test_lmer <- lmer(diff ~ 1 + (1 | new_ID), data = natsap_tidy)
fixef(test_lmer)
ranef(test_lmer)
```

Clean up workspace.

```{r}
cleanup("diff_prog_hier")
rm(some_programs, alphas)
```

```{r}
beep()
```
