---
title: "Stan template"
author: "Sean Raleigh"
date: "July 11, 2015"
output: html_document
---

```{r}
library(knitr)
opts_chunk$set(eval = FALSE)
```

## Using this template

Copy and paste this template to define and run a Stan model.

* Create a new folder in the main working directory called `root`. (Usually `root` is the name of your complete pooling model. E.g., if you are working on a model called `OQdiff_sex_prog_hier`, then your root folder would be `OQdiff_sex`. This will also usually be the name of the `.rmd` file in which you're working.)
* Find and replace all instances of `root`.
* Find and replace `yourmodel` with your model name.
* Add data to the data list defined in `data_yourmodel`.
* Specify the Stan model code.
* Indicate parameters to graph in `params_yourmodel`.
* Change layout (if necessary) in the `sample_plots` function call.
* Add extra variables (if necessary) to `rm`.
* When you have a working model, change `final_model` to TRUE.

## Template:

Gather data to pass to Stan:

```{r}
# model name is 'yourmodel'

data_yourmodel <- list() ##### <----- ADD DATA HERE. #####
```

Write Stan model:

```{r}
cat("
    data {                ##### <----- ADD MODEL HERE. #####
    }
    parameters {
    }
    model {
    }
",
file = "./root/yourmodel.stan")
```

Fit Stan model:

```{r}
final_model = FALSE ##### <----- CHANGE TO TRUE ONCE THE MODEL WORKS. #####

if (file.exists("./root/yourmodel.rds") & final_model == TRUE) {
    stan_yourmodel <- readRDS("./root/yourmodel.rds")
} else {
    stan_yourmodel <- stan_model("./root/yourmodel.stan")
    saveRDS(stan_yourmodel, file = "./root/yourmodel.rds")
}

fit_yourmodel <- sampling(stan_yourmodel, data = data_yourmodel)
```

Extract and save samples:

```{r}
samples_yourmodel <- as.data.frame(fit_yourmodel)

write_csv(samples_yourmodel, path = "./root/yourmodel.csv")

params_yourmodel <- c() ##### <----- ADD PARAMETERS HERE. #####
```

Show chain diagnostics and posterior summaries:

```{r, fig.width = 9, fig.height = 10}
traceplot(fit_yourmodel, pars = params_yourmodel, inc_warmup = FALSE)

print(fit_yourmodel)

len <- length(params_yourmodel)
sample_plots(samples_yourmodel,
             params = params_yourmodel,
             cred_mass = 0.95,
             layout = matrix(c(1:(len + len %% 2)),
                             nrow = (len + len %% 2)/2,
                             ncol = 2))  ##### <----- CHANGE THIS (IF NEEDED). #####
                                         ##### E.g., byrow = TRUE is common    #####
```

Clean up workspace:

```{r, results = 'hide'}
cleanup("yourmodel")
rm(final_model, len) ##### <----- ADD VARIABLES HERE (IF NEEDED). #####
gc()
```

```{r}
beep()
```
