---
title: "NATSAP- OQ diff (somatic) by RTC/WT and sex, plus dosage predictors"
author: "Sean Raleigh"
date: "August 19, 2015"
output: html_document
---

This document contains the analyis of OQ difference from the somatic subscale with RTC/WT and sex (and their interaction) as predictors, along with other group-level dosage predictors.

## Housekeeping

Load necessary libraries and special functions.

```{r, message = FALSE}
library(rstan)
library(parallel)
library(beepr)
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(stringr)
library(tidyr)
library(lme4)
```

```{r}
source("stan_utilities.R")
```

## Import data

```{r}
natsap_tidy <- read_csv("./tidydata/natsap_tidy.csv")
dose_tidy_z <- read_csv("./tidydata/dose_tidy_z.csv")
```


## Try everything related to dosage

Here we model OQdiff using RTCWT and sex along with all other predictors relating to dosage.

Gather data to pass to Stan:

```{r}
# model name is 'OQdiff_som_RTCWT_by_sex_dosage'

RTCWT <- dose_tidy_z$RTCWT[natsap_tidy$new_ID]
RTCWT <- ifelse(RTCWT == "RTC", 0, 1)  # 0 = RTC, 1 = WT

data_OQdiff_som_RTCWT_by_sex_dosage <- list(y = natsap_tidy$OQdiff_som,
                                           n = length(natsap_tidy$OQdiff_som),
                                           J = NROW(dose_tidy_z),
                                           prog = natsap_tidy$new_ID,
                                           RTCWT = RTCWT,
                                           sex = ifelse(natsap_tidy$sex == "MALE", 0, 1),
                                           IT_freq = dose_tidy_z$IT_freq,
                                           GT_freq = dose_tidy_z$GT_freq,
                                           RFT_freq = dose_tidy_z$RFT_freq,
                                           IPFT_freq = dose_tidy_z$IPFT_freq)
```

Mathematical model:

$$ \alpha_{j} \sim N \left( \mu_{\alpha} + \gamma_{ITfreq} ITfreq + \dots +
    \gamma_{IPFTfreq} IPFTfreq, \sigma_{\alpha} \right) $$
$$ y_{i} \sim ~ N\left( \alpha_{j[i]} + \beta_{female} sex_{i} + 
    \beta_{WT} RTCWT_{i} +
    \beta_{WT \cdot female}  \left(RTCWT_{i} \cdot sex_{i} \right), \sigma_{y}\right) $$

Write Stan model:

```{r}
cat("
    data {
        int<lower = 1> n;                       // clients
        vector<lower = -256, upper = 256>[n] y; // OQ diffs
        vector<lower = 0, upper = 1>[n] RTCWT;  // RTC/WT
        vector<lower = 0, upper = 1>[n] sex;    // sex
        int<lower = 1> J;                       // number of programs
        int prog[n];                            // programs
        vector[J] IT_freq;                      // Ind. therapy
        vector[J] GT_freq;                      // Group therapy
        vector[J] RFT_freq;                     // Remote family therapy
        vector[J] IPFT_freq;                    // In-person family therapy
    }
    parameters {
        real mu_a;                // program-level mean
        real<lower = 0> sigma_a;  // program-level sd 
        vector[J] a;              // intercept for each program
        real g_IT_freq;           // Coef. of Ind. therapy
        real g_GT_freq;           // Coef. of Group therapy
        real g_RFT_freq;          // Coef. of Remote family therapy
        real g_IPFT_freq;         // Coef. of In-person family therapy
        real female;              // coefficient of sex
        real WT;                  // coefficient of RTCWT
        real WT_female;           // coefficient of interaction
        real<lower = 0> sigma_y;  // data-level sd
    }
    model {
        // Implicit uniform priors on everrything
        a ~ normal(mu_a + g_IT_freq * IT_freq + g_GT_freq * GT_freq +
                    g_RFT_freq * RFT_freq + g_IPFT_freq * IPFT_freq, sigma_a);
        for(i in 1:n) {
            y[i] ~ normal(a[prog[i]] + female * sex[i] + WT * RTCWT[i] + 
                            WT_female * (RTCWT[i] * sex[i]), sigma_y);
        }
    }
",
file = "./OQdiff_som_RTCWT_by_sex_dosage/OQdiff_som_RTCWT_by_sex_dosage.stan")
```

Fit Stan model:

```{r}
final_model = FALSE

if (file.exists("./OQdiff_som_RTCWT_by_sex_dosage/OQdiff_som_RTCWT_by_sex_dosage.rds") & final_model == TRUE) {
    stan_OQdiff_som_RTCWT_by_sex_dosage <-            readRDS("./OQdiff_som_RTCWT_by_sex_dosage/OQdiff_som_RTCWT_by_sex_dosage.rds")
} else {
    stan_OQdiff_som_RTCWT_by_sex_dosage <- stan_model("./OQdiff_som_RTCWT_by_sex_dosage/OQdiff_som_RTCWT_by_sex_dosage.stan")
    saveRDS(stan_OQdiff_som_RTCWT_by_sex_dosage, file = "./OQdiff_som_RTCWT_by_sex_dosage/OQdiff_som_RTCWT_by_sex_dosage.rds")
}

fit_OQdiff_som_RTCWT_by_sex_dosage <- sampling(stan_OQdiff_som_RTCWT_by_sex_dosage, data = data_OQdiff_som_RTCWT_by_sex_dosage)
```

Extract and save samples:

```{r}
samples_OQdiff_som_RTCWT_by_sex_dosage <- as.data.frame(fit_OQdiff_som_RTCWT_by_sex_dosage)

write_csv(samples_OQdiff_som_RTCWT_by_sex_dosage, path = "./OQdiff_som_RTCWT_by_sex_dosage/OQdiff_som_RTCWT_by_sex_dosage.csv")

# Get 5 random programs
set.seed(1234)
some_programs <- sort(sample(1:data_OQdiff_som_RTCWT_by_sex_dosage$J, 5))

# Display sample sizes for selected programs.
cat("Sample sizes for programs", some_programs, ": ",
    dose_tidy_z$n[dose_tidy_z$new_ID %in% some_programs])

# Gather up parameter names
alphas <- some_programs %>%
    as.character() %>%
    sapply(function(x) {paste("a[", x, "]", sep = "")}) %>%
    unname()


params_OQdiff_som_RTCWT_by_sex_dosage <- c("mu_a", "sigma_a",
                                          # alphas,
                                          "female", "WT", "WT_female",
                                          "g_IT_freq", "g_GT_freq",
                                          "g_RFT_freq", "g_IPFT_freq",
                                          "sigma_y")
```

Show chain diagnostics and posterior summaries:

```{r, fig.width = 9, fig.height = 10}
traceplot(fit_OQdiff_som_RTCWT_by_sex_dosage, pars = params_OQdiff_som_RTCWT_by_sex_dosage, inc_warmup = FALSE)

print(fit_OQdiff_som_RTCWT_by_sex_dosage)

len <- length(params_OQdiff_som_RTCWT_by_sex_dosage)
sample_plots(samples_OQdiff_som_RTCWT_by_sex_dosage,
             params = params_OQdiff_som_RTCWT_by_sex_dosage,
             cred_mass = 0.95,
             layout = matrix(c(1:12),
                             nrow = 5,
                             ncol = 2, byrow = TRUE))
```

Clean up workspace:

```{r, results = 'hide'}
cleanup("OQdiff_som_RTCWT_by_sex_dosage")
rm(final_model, len, RTCWT, some_programs, alphas)
gc()
```
